\begin{abstract}
Scientific data discovery requires search that reasons, not just retrieves. When a climate scientist queries for pressures above 200~kPa, the search system must recognize that ``200000~Pa'' in an HDF5 archive and ``0.2~MPa'' in an S3 dataset are both matches, which requires arithmetic reasoning rather than textual similarity. It must inspect each dataset to determine what metadata it offers, adapt its strategy when one corpus has rich annotations and another has none, and refine its approach when initial results are insufficient. Few current retrieval systems perform these operations in combination.

We present CLIO Search, a retrieval system for scientific data that separates search reasoning from search execution. CLIO implements a four-stage pipeline: Discover (profile each corpus in 14--22\,ms to learn what data and metadata exist), Reason (select retrieval branches based on corpus characteristics), Transform (execute science-aware operators including dimensional conversion and formula normalization as first-class retrieval branches), and Execute (federate across storage backends with iterative multi-hop refinement). The refinement loop is driven by an LLM query rewriter in connected environments or a deterministic fallback for air-gapped HPC deployments; the fallback achieves 96\% of LLM-assisted P@5 quality. CLIO encodes domain reasoning as deterministic operators within an adaptive orchestration framework, complementing learned similarity with algebraic guarantees.

On cross-unit queries, CLIO achieves 0.99~P@5 where the strongest text baseline (BM25) scores 0.40 and the best learned approach (NC-Retriever) achieves 0.16 dense retrieval accuracy; on same-unit control queries, all methods perform comparably (P@5 0.30--0.45), confirming that the gains are specific to cross-unit matching. On 288 real NOAA observatory documents, precision improves by 87\% over hybrid baselines. At distributed scale on 2.5M arXiv abstracts across 4~NVIDIA GH200 nodes on NCSA DeltaAI, cross-unit queries in Pa, kPa, bar, and atm return identical results at sub-50\,ms latency with 0.94 weak scaling efficiency. Integration with IOWarp blob storage demonstrates selective retrieval (0.5\% corpus inspection) where the native API returns all blobs unfiltered. All code and benchmarks are open-source.
\end{abstract}

%% ===================================================================
