\section{Introduction}

A scientist at a national laboratory runs a combustion simulation that produces pressure measurements at 200~kPa. Three months later, a colleague at a different facility searches for similar simulations. The data exists, stored as ``200000~Pa'' in an HDF5 archive on a parallel filesystem. The search returns nothing, because the retrieval system cannot multiply 200 by 1000. The simulation is re-run, wasting compute allocations. This is not a hypothetical scenario: NASA's \$327M Mars Climate Orbiter was lost because one system used pound-seconds while another used newton-seconds.

Such failures repeat across every scientific domain where measurements are expressed in different units, formats, or storage systems. A commonly cited estimate holds that 60 to 80\% of scientific data is never reused, not for lack of access, but for lack of search~\cite{baker2016reproducibility}. The FAIR data principles, now mandated by NSF, NIH, and DOE, require that data be Findable, Accessible, Interoperable, and Reusable. Automated FAIR assessments show that no major repository achieves even 50\% compliance on reusability. A central bottleneck is search.

This bottleneck is becoming more consequential as AI agents enter scientific workflows. Coscientist~\cite{coscientist2023} autonomously executes wet-lab synthesis. AI Scientist v2~\cite{aiscientistv2_2025} generates research end-to-end. Autonomous agents at ORNL orchestrate experiments across national facilities~\cite{ornl_agents_sc25}. These agents can reason about science, plan multi-step experiments, and call external tools. What they cannot do is search scientific data effectively. ScienceAgentBench~\cite{scienceagentbench2025} reports 32.4\% end-to-end task success, with failures traced to data handling rather than reasoning. A comprehensive survey~\cite{agentic_ai_survey2025} identifies scientific data management as explicitly underexplored in the agent literature. The retrieval infrastructure these agents depend on treats all content as plain text, inheriting the same failures that prevent human scientists from finding existing data.

Current retrieval systems fall short of scientific data discovery in three fundamental ways.

First, scientific search requires domain reasoning that retrieval systems do not perform. A query for pressures between 190 and 360~kPa must recognize that ``200000~Pa'' and ``0.2~MPa'' are both matches, but this requires arithmetic, not text similarity. Formulas compound the problem: ``$F = ma$,'' ``$F = m \cdot a$,'' and ``$ma = F$'' express the same physical law but produce distinct token sequences that defeat both keyword and embedding retrieval. The Numeracy Gap study~\cite{numeracygap2026} evaluates 13 embedding models and reports 0.54 binary retrieval accuracy on numerical content. String normalization~\cite{numbersmatter2024} cannot bridge SI prefixes. Learned approaches~\cite{cone2026,ncretriever2025} achieve up to 65\% recall improvement but provide no correctness guarantee.

Second, scientific datasets vary enormously in metadata quality, and no search system adapts to this variation. Automated FAIR metadata assessments consistently find that fewer than half of datasets in major repositories carry structured, machine-readable annotations~\cite{baker2016reproducibility}. An HDF5 file with Climate and Forecast convention attributes provides explicit units for every variable; a raw binary dump on a parallel filesystem may offer nothing beyond a filename. A fixed retrieval pipeline cannot handle both: it either wastes computation on datasets with no extractable measurements or misses opportunities when rich metadata is available. A search system that profiles each corpus before querying, determining what metadata exists and which strategies are viable, would address this gap, but no current retrieval system provides this capability.

Third, scientific data is distributed across storage systems that no search engine spans with domain awareness. A single experimental campaign may deposit raw output on a parallel filesystem, preprocessed arrays in HDF5, derived features in a vector database, and catalog entries on an external platform. Existing systems either federate generic text search without science-aware operators~\cite{ragroute2025}, or capture storage provenance without exposing a search interface~\cite{provio2024}. No existing system combines all three: domain-aware operators, corpus-adaptive orchestration, and federated HPC access.

These challenges share a common requirement: the search system must reason about \emph{how} to search before it executes the search. It must observe what each corpus contains, decide which strategies to apply, act with domain-specific operators, and evaluate whether to refine. This observe-decide-act-evaluate cycle is what separates agentic search from conventional retrieval, and it is precisely what current systems lack.

We present CLIO Search, a retrieval system that implements this cycle for scientific data. CLIO profiles each corpus to discover what metadata exists (completing in 14--22\,ms, scaling sub-linearly to thousands of namespaces via SQL aggregation on indexed columns), selects retrieval branches based on corpus characteristics (skipping branches that cannot contribute), executes science-aware operators as deterministic database predicates complementing BM25 and dense vector search, and federates across heterogeneous storage backends. An iterative refinement loop expands, narrows, or pivots the query based on intermediate results; the loop is driven by an LLM query rewriter in connected environments or a deterministic fallback for air-gapped HPC deployments, where the fallback achieves 96\% of LLM-assisted P@5 quality. Domain knowledge enters the pipeline as executable, verifiable computation: the fact that 5~km = 5000~m is an algebraic invariant encoded in the retrieval operator, guaranteed correct for all supported unit transformations. Cross-unit queries constitute a significant fraction of scientific data discovery: pressure alone is expressed in seven incompatible units across disciplines (Pa, hPa, kPa, MPa, bar, atm, psi), temperature in three scales (K, \textdegree C, \textdegree F), and energy in seven variants (eV, keV, MeV, GeV, kJ, MJ, GJ).

We make three contributions:

\begin{enumerate}
\item \textbf{Science-aware retrieval operators.} We introduce dimensional conversion and formula normalization (algebraic equivalence matching, e.g., $F{=}ma$ vs.\ $ma{=}F$) as deterministic, first-class retrieval branches that execute alongside BM25 and dense vector search. A composable unit registry based on dimensional analysis canonicalizes measurements to SI base units at index time, enabling cross-unit queries as database range predicates with correctness guaranteed for all supported unit transformations. The registry covers 46~units across 12~physical domains\footnote{Length, mass, time, temperature, pressure, velocity, frequency, radioactivity, energy, power, area, and acceleration.}; adding a new unit requires one configuration entry, not code changes.

\item \textbf{Corpus-adaptive search orchestration.} A profiling layer inspects each namespace before querying, learning what metadata exists and which retrieval branches are viable. The system adapts its search strategy per dataset: rich-metadata corpora activate science-aware operators, sparse corpora fall back to text-based retrieval, and multi-hop refinement expands, narrows, or pivots the query based on intermediate results.

\item \textbf{Federated scientific search.} A connector architecture provides unified search across heterogeneous storage backends (filesystems, object stores, HDF5, NetCDF, IOWarp blob storage, external catalogs) with per-connector capability negotiation. Each connector declares which retrieval branches it supports, and the coordinator routes queries accordingly.
\end{enumerate}

We evaluate on a controlled benchmark (210~documents, 80~queries), real observatory data (288~NOAA GHCN-Daily documents), a live external catalog (341~NDP datasets), distributed HPC infrastructure (2.5M arXiv abstracts across 4~GH200 nodes on NCSA DeltaAI), and IOWarp blob storage (10K blobs). CLIO achieves 0.99~P@5 on cross-unit queries (vs.\ 0.40 for the strongest text baseline); on same-unit controls all methods score comparably (0.30--0.45), confirming the gains are specific to cross-unit matching. Real-world precision improves by 87\% on NOAA data, with sub-50\,ms query latency and 0.94 weak scaling efficiency at distributed scale.

%% ===================================================================
%% Section 2: Background and Related Work
